---
output: github_document
editor_options: 
  chunk_output_type: console
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "images/"
)

##version <- as.vector(read.dcf('DESCRIPTION')[, 'Version'])
##version <- gsub('-', '.', version)
version <- "0.0.9000"
```

# gnumaker

Version: `r version`

## Overview

**gnumaker** makes if easy to create and use GNU Makefiles to aid a reproducible work flow for data analysis projects.

GNU Make is the defacto standard for efficiently rerunning appropriate steps in the data analysis or reporting process if a particular file is changed. Only the necessary steps are rerun.

Rather than creating a new system for setting up and building output from statistical software syntax files, **gnumaker** leverages off existing GNU Make rules. These rules, for R, Sweave, R Markdown, Stata, SAS and other syntax files are available at [r-makefile-definitions on Github](https://github.com/petebaker/r-makefile-definitions). These are described in P Baker (2019) Using GNU Make to Manage the Workflow of Data Analysis Projects, _Journal of Statistical Software (Accepted)_. 

For those not familiar with GNU Make, **gnumaker** allows simple dependencies between files to be specified to produce a working Makefile and the associated directed acyclic graph (DAG). 
I'd welcome Github issues containing error reports or feature requests. 
Alternatively, you can email the package maintainer at drpetebaker at gmail dot com.

## Installation

<!--
Install the latest CRAN version of **gnumaker** with:

```{r cran-installation, eval = FALSE}
##install.packages("gnumaker")
```
-->

You can install the development version of **gnumaker** from GitHub with:
```{r gh-installation, eval = FALSE}
## if you don't have devtools installed, first run:
## install.packages("devtools")
devtools::install_github("petebaker/gnumaker")
```

## Usage

There are three key functions in **gnumaker**. These are:

* `create_makefile()` creates a gnu\_makefile object given dependencies between syntax, data and output files,
* `plot_makefile()` plots a DAG for a gnu\_makefile object and
* `write_makefile()` writes a Makfile to disk.

## Examples

Suppose we have a data file `simple.csv` and use `read.R` to read and
clean the data. After storing the cleaned data in a .RData file, we
then employ `linmod.R` to plot and analyse the data. Next, using the
stored results two reports, `report1.pdf` and `report2.docx` are
produced from `report1.Rmd` and `report2.Rmd`. The workflow may be
encapsulated in a Makefile which is then employed to manage the
process and generate or regenerate any intermediate files when the
data or syntax changes.

```{r, eval = FALSE, simple-demo}
library(gnumaker)

## Here we assume the default extension for running R file is Rout
## (default) but instead specify the output types for the .Rmd files
## targets 'read' depends on read.r & simple.csv, 'linmod' on linmod.R
## and 'read' etc
gm1 <-
  create_makefile(targets = list(read = c("read.R", "simple.csv"),
                                 linmod = c("linmod.R", "read"),
                                 rep1 = c("report1.Rmd", "linmod"),
                                 rep2 = c("report2.Rmd", "linmod")),
                  phony = list(all = c("rep1", "rep2")),
                  comments = list(linmod = "plots and analysis using 'linmod.R'")
                  file.exts = list(Rmd = list(rep1 = "pdf", rep2 = "docx")))
write_makefile(gm1)
plot_makefile(gm1)                  
```

The Makefile produced with `write_makefile(gm1)` is

```{bash, eval = FALSE, simple-makefile}
.PHONY: all
all: report1.pdf report2.docx

# produce report(s) from .Rmd: depends on linmod.Rout
report1.pdf: ${@:.pdf=.Rmd} linmod.Rout

# produce report(s) from .Rmd: depends on linmod.Rout
report2.docx: ${@:.docx=.Rmd} linmod.Rout

# plots and analysis using 'linmod.R'
linmod.Rout: ${@:.Rout=.R} read.Rout

# read.Rout: depends on read.R and simple.csv
read.Rout: ${@:.Rout=.R} simple.csv

include r-rules.mk
```

The DAG of the Makefile produced with `plot_makefile(gm1)` is

![DAG for simple demo Makefile. Using the minimal set of files (shown
in green rectangles), then GNU Make allows us to (re)generate all
other files shown as wheat coloured
circles))](images/simple-demo.png)


For more examples, see the gnumaker vignette (under construction).
