---
output: github_document
editor_options: 
  chunk_output_type: console
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "images/"
)

version <- as.vector(read.dcf('DESCRIPTION')[, 'Version'])
version <- gsub('-', '.', version)
##version <- "0.0.900X"
```

# gnumaker

Version: `r version`

## Overview

**gnumaker** makes if easy to create and use GNU Makefiles to aid a reproducible work flow for data analysis projects.

GNU Make is the defacto standard for efficiently rerunning appropriate steps in the data analysis or reporting process if a particular file is changed. Only the necessary steps are rerun.

Rather than creating a new system for setting up and building output from statistical software syntax files, **gnumaker** leverages off existing GNU Make rules. These rules, for R, Sweave, R Markdown, Stata, SAS and other syntax files are available at [r-makefile-definitions on Github](https://github.com/petebaker/r-makefile-definitions). These are described in P Baker (2019) Using GNU Make to Manage the Workflow of Data Analysis Projects, _Journal of Statistical Software (Accepted)_. 

For those not familiar with GNU Make, **gnumaker** allows simple dependencies between files to be specified to produce a working Makefile and the associated directed acyclic graph (DAG). 
I'd welcome Github issues containing error reports or feature requests. 
Alternatively, you can email the package maintainer at drpetebaker at gmail dot com.

## Installation

<!--
Install the latest CRAN version of **gnumaker** with:

```{r cran-installation, eval = FALSE}
##install.packages("gnumaker")
```
-->

You can install the development version of **gnumaker** from GitHub with:
```{r gh-installation, eval = FALSE}
## if you don't have devtools installed, first run:
## install.packages("devtools")
devtools::install_github("petebaker/gnumaker")
```

## Usage

There are three key functions in **gnumaker**. These are:

* `create_makefile()` creates a gnu\_makefile object given dependencies between syntax, data and output files,
* `write_makefile()` writes a Makfile to disk.
* `plot()` plots a DAG for a gnu\_makefile object and

## Examples

Suppose we have a data file `simple.csv` and use `read.R` to read and
clean the data. After storing the cleaned data in a .RData file, we
then employ `linmod.R` to plot and analyse the data. Next, using the
stored results, two reports `report1.pdf` and `report2.docx` are
produced from `report1.Rmd` and `report2.Rmd`. The workflow may be
encapsulated in a Makefile which is then employed to manage the
process and generate or regenerate any intermediate files when the
data or syntax changes.

Using the **gnumaker** package we simply need to provide a list of
targets to the the `create_makefile` function where the components
specify a target and dependency file(s). The package uses the GNU Make
pattern rules in _r-rules.mk_ to choose file names for targets but we
can override the defaults. For instance, in the example below we
provide the first target as the first component of the list as 
`read = c("read.R", "simple.csv")` the second target depends on the `read`
target and `linmod.R` and so we specify this with 
`linmod = c("linmod.R", "read")` and so on. 

Target file names are substituted using defaults and the _Makefile_ is
rearranged using the DAG of the relationships. For instance, the
default target file for the first dependency in the `read` component,
which is `read.R`, becomes `read.Rout` but we can change the default
target file extension for all `.R` files using the `default.exts`
argument and specify say a HTML target file with 
`default.exts = list(R = "html")`.


```{r, simple-demo}
library(gnumaker)
gm1 <-
  create_makefile(targets = list(read = c("read.R", "simple.csv"),
                  linmod = c("linmod.R", "read"),
                  rep1 = c("report1.Rmd", "linmod"),
                  rep2 = c("report2.Rmd", "linmod")),
                  target.all = c("rep1", "rep2"),
                  all.exts = list(rep1 = "pdf", rep2 = "docx"),
 comments = list(linmod = "plots and analysis using 'linmod.R'"))
```

A Makefile `Makefile.demo` is produced with `write_makefile(gm1)`

```{r, simple-makefile}
write_makefile(gm1, file = "Makefile.demo")
```

```{bash, simple-makefile-gm1, comment="", echo = FALSE}
cat Makefile.demo
```

The DAG of the `gnu_makefile` object can be produced with `plot(gm1)`.

```{r, simple-dag, fig.cap = "DAG of Makefile for simple example. The DAG of the `gnu_makefile` object can be produced with `plot(gm1)`. Using the minimal set of files (shown in green rectangles), then GNU Make allows us to (re)generate all other files shown as wheat coloured circles)"}
plot(gm1)
```

For more examples, see the gnumaker vignette (under construction).

## Notes

**gnumaker** is under construction and should change (and improve)
rapidly over the next few months.

## To do

- DONE extract dependency and target file extensions in `r-rules.mk`,
  preferably by parsing the included file (done using `pattern-exts`')
- DONE incorporate dependency and target file extensions extracted
  using `pattern-exts` into `create_makefile` and set defaults
- TODO move `pattern_exts` to internal functions and create
  `show_extensions` to assist user specification
- TODO allow specification of _global options_ in `zzz.R` so that it
  is easier to customise defaults e.g. so user can specify defaults in
  .Rprofile
- TODO add `testthat` unit testing for more complicated examples
- TODO add travis.ci and other automatic checking - see [r-pkgs.had.co.nz/release.html](http://r-pkgs.had.co.nz/release.html)
- TODO allow for target file extensions and dependency files to be set
  as user specified variables which would make the `Makefile`s
  produced more flexible but less easy to read
- TODO allow `target.all` to be determined from DAG if this is sensible
- TODO either incorporate `makefile2graph` as a way of plotting `Makefile`s not
  made with **gnumaker** or write own functions.
  (See [makefile2graph on github](https://github.com/lindenb/makefile2graph "makefile2graph on github"))
