---
output: github_document
editor_options: 
  chunk_output_type: console
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "images/"
)

version <- as.vector(read.dcf('DESCRIPTION')[, 'Version'])
version <- gsub('-', '.', version)
##version <- "0.0.900X"
```

# gnumaker

Version: `r version`

## Overview

**gnumaker** makes it easy to create and use GNU Makefiles to aid a reproducible work flow for data analysis projects.

GNU Make is the defacto standard for efficiently rerunning appropriate steps in the data analysis or reporting process if a particular file is changed. Only the necessary steps are rerun.

Rather than creating a new system for setting up and building output
from statistical software syntax files, **gnumaker** leverages off
existing GNU Make rules. These rules, for `R`, `Sweave`, `R Markdown`,
`Stata`, `SAS`, `PSPP`, `Python`, `Perl` and other syntax files are
available at [r-makefile-definitions on
Github](https://github.com/petebaker/r-makefile-definitions). These
are described in P Baker (2020) Using `GNU Make` to Manage the Workflow
of Data Analysis Projects, _Journal of Statistical Software
(Accepted)_.

For those not familiar with `GNU Make`, **gnumaker** allows simple
dependencies between files to be specified to produce a working
`Makefile` and the associated directed acyclic graph (DAG).  I'd
welcome `Github` issues containing error reports or feature requests.
Alternatively, you can email the package maintainer at `drpetebaker at
gmail dot com`.

## Installation

<!--
Install the latest CRAN version of **gnumaker** with:

```{r cran-installation, eval = FALSE}
##install.packages("gnumaker")
```
## Note that three dependencies are in BioConductor so use BiocManager
NB: parked here in case biocViews: line in DESCRIPTION does not work

if (!requireNamespace("BiocManager", quietly = TRUE))
   install.packages("BiocManager")
devtools::install_github("petebaker/gnumaker", repos = BiocManager::repositories())
-->

You can install the development version of **gnumaker** from `GitHub` with:
```{r gh-installation, eval = FALSE}
## if you don't have devtools installed, automatically install it from CRAN
if (!requireNamespace("devtools", quietly = TRUE))
   install.packages("devtools")
devtools::install_github("petebaker/gnumaker")
```

## Usage

There are currently four key functions in **gnumaker**. These are:

* `create_makefile()` creates a `gnu_makefile` object given the
  specified dependencies between syntax, data and output files,
* `write_makefile()` writes a `gnu_makefile` object to a `Makfile` on disk,
* `info_rules()` provides information about data analysis `GNU Make` rules for various target and dependency filename extensions, and
* `plot()` plots the DAG of a `gnu_makefile` object.

## Example

Suppose we have a data file `simple.csv` and use `read.R` to read and
clean the data. After storing the cleaned data in a .RData file, we
then employ `linmod.R` to plot and analyse the data. Next, using the
stored results, two reports `report1.pdf` and `report2.docx` are
produced from `report1.Rmd` and `report2.Rmd`. The workflow may be
encapsulated in a `Makefile` which is then employed to manage the
process and generate or regenerate any intermediate files when the
data or syntax changes. 

Rather than writing `GNU Make` commands, we can employ the
**gnumaker** package to create an appropriate `Makefile` called
`Makefile.demo` with:

```{r, simple-demo0}
library(gnumaker)
gm1 <-
  create_makefile(targets = list(read = c("read.R", "simple.csv"),
                  linmod = c("linmod.R", "read"),
                  rep1 = c("report1.Rmd", "linmod"),
                  rep2 = c("report2.Rmd", "linmod")),
                  target.all = c("rep1", "rep2"),
                  all.exts = list(rep1 = "pdf", rep2 = "docx"))
write_makefile(gm1, file = "Makefile.demo")
```

To plot the corresponding DAG, which shows the relationships between
target files (wheat coloured circles) which we can (re)generate from a
minimal set of syntax and data files (green rectangles) using `GNU Make`, use:

```{r, simple-dag, fig.cap = "DAG of `Makefile` for simple example. The DAG of the `gnu_makefile` object can be produced with `plot(gm1)`. Using the minimal set of files (shown in green rectangles), then `GNU Make` allows us to (re)generate all other files shown as wheat coloured circles)"}
plot(gm1)
```

### Details

Using the **gnumaker** package we simply need to provide a list of
targets to the the `create_makefile` function where the components
specify a target as a name and dependency file(s) as a character
vector. The package uses the `GNU Make` pattern rules in `r-rules.mk`
to choose file names for targets but we can override the defaults.

For instance, in this example the first two dependency files are
`simple.csv` and `read.R` so we provide the first target as the first
component of the list as `read = c("read.R", "simple.csv")`, where the
name `read` can be anything we like. Note that the target filename for
`read` will be derived from the first filename in the list `read.R`.
The second target depends on the `read` target and `linmod.R` and so
we specify this with `linmod = c("linmod.R", "read")` and so on.

Using the function `create_makefile` to create a `gnu_makefile`
object, target file names are substituted using defaults and the
appropriate `Make` commands are rearranged using the DAG of the
relationships. For instance, the default target file for the first
dependency in the `read` component, which is `read.R`, becomes
`read.Rout` but we can change the default target file extension for
all `.R` files using the `default.exts` argument and specify say a
HTML target file with `default.exts = list(R = "html")`.

Finally we specify the first target (usually `all`) as two reports
`report1.pdf` and `report2.docx` using `target.all = c("rep1","rep2")`
which by default would be `report1.html` and `report2.html` but which
we specify as `report1.pdf` and `report2.docx` by specifying the
option `all.exts = list(rep1 = "pdf", rep2 = "docx")`.
 
Once we have constructed a suitable `gnu_makefile` object then we
write it to disk with `write_makefile`. To run all `R` script files
and analyses in order we simply type `make` in a terminal or set up
`RStudio` or our IDE to use GNU Make as the build mechanism which
allows us to (re)run analyses by pressing the appropriate Build
button.
 
The `Makefile`, with optional comment for `linmod` target, is
specified using:

```{r, simple-demo1}
library(gnumaker)
gm1 <-
  create_makefile(targets = list(read = c("read.R", "simple.csv"),
                  linmod = c("linmod.R", "read"),
                  rep1 = c("report1.Rmd", "linmod"),
                  rep2 = c("report2.Rmd", "linmod")),
                  target.all = c("rep1", "rep2"),
                  all.exts = list(rep1 = "pdf", rep2 = "docx"),
 comments = list(linmod = "plots and analysis using 'linmod.R'"))
```

A Makefile `Makefile.demo` is produced with `write_makefile(gm1)`

```{r, simple-makefile1}
write_makefile(gm1, file = "Makefile.demo")
```

```{bash, simple-makefile-gm1, comment="", echo = FALSE}
cat Makefile.demo
```

<!-- The DAG of the `gnu_makefile` object can be produced with `plot(gm1)`. -->

<!-- ```{r, simple-dag2, fig.cap = "DAG of Makefile for simple example. The DAG of the `gnu_makefile` object can be produced with `plot(gm1)`. Using the minimal set of files (shown in green rectangles), then GNU Make allows us to (re)generate all other files shown as wheat coloured circles)"} -->
<!-- plot(gm1) -->
<!-- ``` -->

We can use the function `info_rules` to determine the possible target
files for dependency files. For instance, what target files have
`Makefile` rules for an `.R` `R` syntax file?

```{r, inforulesr}
info_rules("R")
```

For `.Rmd` R Markdown files, use

```{r, inforulesrmd}
info_rules("Rmd")
```

For more examples, see the **gnumaker** vignette (under construction).

## Note

The **gnumaker** `R` package is under construction and could change
(and improve) rapidly at various times but this depends on work/life
balance.
